---
title: "DSE 511 Final Project: Data Cleaning"
author: "Cecilia Albert-Black"
date: "2025-11-04"
output: html_document
---

Load necessary libraries

```{r}
library(readr)
library(dplyr)
library(stringr)
```

## Cleaning the NORS Outbreak Dataset

Set working directory to repo, then import raw dataset

```{r}
# Import
outbreak_df <- read_csv("NORS_outbreaks.csv", show_col_types = FALSE)

# Preview
head(outbreak_df)
```

### Understanding/Inspecting the Dataset

Identify columns and their structure

```{r}
# Structure
str(outbreak_df)
```

Summary statistics by column using skimr package

```{r}
library(skimr)
skim(outbreak_df)
```

Determine unique values of categorical variables

1.  States

```{r}
unique(outbreak_df$State)
```

2.  Primary Mode

```{r}
# Table of unique primary modes
table(outbreak_df$`Primary Mode`)
```

3.  Etiology

```{r}
# Number of unique etiologies
n_distinct(outbreak_df$Etiology)
```

Many different etiologies, many with similar words. I will use dplyr to detect keywords and assign them to broader categories within a new column called Etiology_grouped to decrease the number of unique etiologies (may be helpful to data analysis later on):

```{r}
outbreak_df <- outbreak_df %>%
  mutate(Etiology_grouped = case_when(
    str_detect(Etiology, regex("norovirus", ignore_case = TRUE)) ~ "Norovirus",
    str_detect(Etiology, regex("salmonella", ignore_case = TRUE)) ~ "Salmonella",
    str_detect(Etiology, regex("campylobacter", ignore_case = TRUE)) ~ "Campylobacter",
    str_detect(Etiology, regex("clostridium", ignore_case = TRUE)) ~ "Clostridium",
    str_detect(Etiology, regex("staphylococcus", ignore_case = TRUE)) ~ "Staphylococcus",
    str_detect(Etiology, regex("hepatitis", ignore_case = TRUE)) ~ "Hepatitis",
    str_detect(Etiology, regex("unknown|unspecified|not determined", ignore_case = TRUE)) ~ "Unknown",
    TRUE ~ "Other"
  ))

# Check new grouping
n_distinct(outbreak_df$Etiology_grouped)
```

Top etiology groups

```{r}
outbreak_df %>%
  count(`Food Vehicle`, sort = TRUE)
```

3.  Serotype or Genotype

```{r}
# Number of unique variables
n_distinct(outbreak_df$`Serotype or Genotype`)
```

We'll keep this variable as-is since it describes in more detail the etiology of the outbreak

3.  Etiology Status

```{r}
# Number of unique variables
n_unique(outbreak_df$`Etiology Status`)
```

```{r}
# Unique names
unique(outbreak_df$`Etiology Status`)
```

Very messy variable that is hard to interpret - maybe we'll omit it or find a source that describes it better.

3.  Setting

```{r}
n_unique(outbreak_df$Setting)
```

Many different settings, some have similarities, might be worth exploring ways to group them.

3.  Food Vehicle

```{r}
# Top 10 food vehicles by rank
outbreak_df %>%
  count(`Food Vehicle`, sort = TRUE) %>%
  slice_head(n = 10)
```

Investigate where the missing values of food vehicle are coming from

```{r}
outbreak_df %>%
  mutate(FoodVehicle_missing = if_else(is.na(`Food Vehicle`) | `Food Vehicle` == "", "Missing", "Present")) %>%
  count(`Primary Mode`, FoodVehicle_missing) %>%
  group_by(`Primary Mode`) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()

```

This table shows that about half of the reported food vehicles are missing within the category food within primary mode.

3.  Food Contaminated Ingredient

```{r}
# Number of unique variables 
n_unique(outbreak_df$`Food Contaminated Ingredient`)
```

This might be a redundant variable, consider omitting and just using food vehicle for sake of simplicity.

3.  IFSAC Category

IFSAC stands for the Interagency Food Safety Analytics Collaboration, which is a categorization scheme used to identify foods most often linked to specific illnesses caused by certain pathogens. Source: <https://www.cdc.gov/ifsac/php/projects/food-categorization-scheme.html#:~:text=At%20a%20glance,include%20ice%20and%20dietary%20supplements.>

```{r}
unique(outbreak_df$`IFSAC Category`)
```

This category is similar to the food vehicle category; because it's more standardized, it might be worth using instead of food category.

3.  Water Exposure

```{r}
# Top 10 water exposure types by rank
outbreak_df %>%
  count(`Water Exposure`, sort = TRUE)
```

Investigate where the missing values are coming from

```{r}
outbreak_df %>%
  mutate(Waterexposure_missing = if_else(is.na(`Water Exposure`) | `Water Exposure` == "", "Missing", "Present")) %>%
  count(`Primary Mode`, Waterexposure_missing) %>%
  group_by(`Primary Mode`) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()

```

All outbreaks within water as primary mode of exposure have a designated type of water exposure (in other words, no missing values of water exposure in water-borne outbreaks)

3.  Water Type

```{r}
n_unique(outbreak_df$`Water Type`)
```

```{r}
# Top 10 water types by rank
outbreak_df %>%
  count(`Water Type`, sort = TRUE)
```

Some water types have similar names, so I will create a grouped variable to lump some of them into similar categories

```{r}
library(dplyr)
library(stringr)

outbreak_df <- outbreak_df %>%
  mutate(Water_type_group = case_when(
    is.na(`Water Type`) ~ "Missing/NA",
    str_detect(`Water Type`, regex("community|individual|private", ignore_case = TRUE)) ~ "Drinking/Community Water",
    str_detect(`Water Type`, regex("pool|hot tub|spa|whirlpool|splash pad|fountain|water playground", ignore_case = TRUE)) ~ "Recreational Water",
    str_detect(`Water Type`, regex("lake|reservoir|other", ignore_case = TRUE)) ~ "Natural/Other",
    str_detect(`Water Type`, regex("unknown|unspecified|not determined", ignore_case = TRUE)) ~ "Unknown",
    TRUE ~ "Other"
  ))

# Check the new grouping
outbreak_df %>%
  count(Water_type_group, sort = TRUE)

```

3.  Animal Type

```{r}
# Top 10 animal types by rank
outbreak_df %>%
  count(`Animal Type`, sort = TRUE)
```

Create a new category which summarizes the animal type exposures into groups

```{r}
library(dplyr)
library(stringr)

outbreak_df <- outbreak_df %>%
  mutate(Animal_group = case_when(
    is.na(`Animal Type`) ~ "Missing/NA",
    str_detect(`Animal Type`, regex("poultry", ignore_case = TRUE)) ~ "Poultry",
    str_detect(`Animal Type`, regex("cattle", ignore_case = TRUE)) ~ "Cattle",
    str_detect(`Animal Type`, regex("cat|dog|ferret|guinea pig|hedgehog|mouse|rat|raccoon", ignore_case = TRUE)) ~ "House Pets",
    str_detect(`Animal Type`, regex("horse|pony|goat|sheep|pig|alpaca|donkey|llama|yak", ignore_case = TRUE)) ~ "Other Ungulates",
    str_detect(`Animal Type`, regex("lizard|turtle|snake|reptile", ignore_case = TRUE)) ~ "Reptiles",
    str_detect(`Animal Type`, regex("bird, not including poultry", ignore_case = TRUE)) ~ "Other Birds",
    str_detect(`Animal Type`, regex("fish", ignore_case = TRUE)) ~ "Fish",
    str_detect(`Animal Type`, regex("frog|salamander", ignore_case = TRUE)) ~ "Amphibians",
    str_detect(`Animal Type`, regex("other", ignore_case = TRUE)) ~ "Other/Unknown",
    TRUE ~ "Other/Unknown"
  ))

# Check the counts for each group
outbreak_df %>%
  count(Animal_group, sort = TRUE)


```
