---
title: "DSE 511 Final Project Cluster Analysis in R"
author: "Cecilia Albert-Black"
date: "2025-11-11"
output: html_document
---

Set working directory to R code folder

```{r}
setwd("C:/Users/q4q/repos/DSE511_Final_Project/Step 3. Cluster Analysis/R Code")
```

Load necessary libraries (must be installed first)

```{r}
library(readr)
library(tidyverse)
library(fitdistrplus)
library(factoextra) #for pca
library(klaR) # K-modes clustering
library(psych) # for KMO test
```

Import dataset from Step 1. Data Cleaning

```{r}
clean_df <- read_csv("C:/Users/q4q/repos/DSE511_Final_Project/Step 1. Data Cleaning/R code/outbreak_clean_r_v3.csv")

head(clean_df)
```

## Cluster Analysis

### I. Hierarchical

**Assumptions**

-   Data must be numeric or convertible to a distance metric

-   Variables must be on comparable scales (standardized variables)

-   Data should have some natural separation, the structure cannot be completely random

-   Variables can be non-parametric

**Limitations encountered:** hclust() function in base R could not handle more than 65,536 observations (the dataset has 66,713).

### II. K-Means

**Assumptions**: continuous variables with Euclidean distances.

**Limitations**: our data fail these assumptions because they are discrete, right-skewed, and Poisson/logarithmically distributed. Thus, we cannot perform a k-means cluster analysis.

**Notes**: I explored this option regardless of failing the assumptions in the DSE511_Final_Project_Cluster_Analysis_V3.

### III. K-Modes

**Assumptions**

-   Variables must be categorical

-   Variables must not have any missing data

-   Categories must have a mode (most frequent category) as their center

-   Categories must have a similar size/density

**Limitations**: None!

**Notes**: some categorical variables only apply to some of the observations; thus, we will replace NAs with "Not Applicable".

Select categorical variables that applt to all outbreaks by creating a vector of character names

```{r}
cat_vars <- c(
  "Date",
  "Season",
  "Region",
  "Primary_mode",
  "Setting_group",
  "Etiology_group"
)
```

Create a categorical dataset and convert to factors

```{r}
kmodes_df <- clean_df %>%
  dplyr::select(all_of(cat_vars)) %>%
  mutate(
    across(everything(), as.factor)
  )

# sanity check
dim(kmodes_df)
names(kmodes_df)
```

Ensure there are no missing values in these categories (there should be no TRUE counts)

```{r}
summary(is.na(kmodes_df))
```

Convert to dataframe

```{r}
kmodes_input <- as.data.frame(kmodes_df)
```

#### i. Determining number of clusters (Elbow method)

Run k-modes on different sizes (e.g. from 1-10)

```{r}
set.seed(123)

kmodes_results <- vector("list", length = 10)

for (k in 1:10) {
  cat("Fitting k =", k, "\n")
  
  kmodes_results[[k]] <- klaR::kmodes(
    kmodes_input,
    modes   = k,
    iter.max = 50
  )
}

names(kmodes_results) <- paste0("k_", 1:10)
```

For each size, plot the cost function (i.e. within-cluster dissimilarities) against the number of clusters (k) to identify the elbow point

```{r}
# Extract within-cluster dissimilarities (cost) by cluster size
kmodes_costs <- sapply(
  kmodes_results,
  function(x) sum(as.numeric(x$withindiff))
)

plot(
  1:10, kmodes_costs, type = "b",
  xlab = "Number of clusters (k)",
  ylab = "Within-cluster dissimilarity (cost)",
  main = "k-modes cost vs number of clusters"
)
```

Identify the elbow point in the plot, where the rate of decrease in the cost function significanlty slows down; this point suggests the optimal number of clusters.

It seems like 2 clusters could be sufficient but 2-6 clusters have a similar cost.

Since it's hard to determine what cluster size to use, let's narrow our results by omitting clusters with very small sizes

#### ii. Assessing cluster sizes

```{r}
# Extract cluster size tables
cluster_sizes <- lapply(kmodes_results, function(x) table(x$cluster))

# Plot each one separately
par(mfrow = c(2, 2))  # adjust if you want a grid of plots

for (k in 1:10) {
  barplot(
    cluster_sizes[[k]],
    main = paste("Cluster Sizes for k =", k),
    xlab = "Cluster",
    ylab = "Count",
    col = "skyblue"
  )
}
```

We'll land on 4 clusters since it seems to have roughly equal cluster sizes while not over clustering.

#### iii. Results of 2 Clusters

```{r}
# 2 clusters 
kmodes_result_2 <- klaR::kmodes( 
  kmodes_input, 
  modes = 2, 
  iter.max = 50 
  )
```

Once you determined the optimal number of clusters, determine:

Size of each cluster

```{r}
table(kmodes_result_2$cluster)
```

Most common categorizes per cluster

```{r}
kmodes_result_2$modes
```

Add clusters back to dataset

```{r}
# Add cluster labels to the kmodes_input data
kmodes_clusters_2 <- factor(kmodes_result_2$cluster)

kmodes_df_2 <- kmodes_input %>%
  mutate(cluster = kmodes_clusters_2)

# Quick look at cluster sizes
kmodes_df_2 %>%
  count(cluster)
```

Visualize

```{r}
library(ggplot2)
library(dplyr)

# Helper function
plot_cluster_comp <- function(data, var_name) {
  ggplot(
    data,
    aes(x = .data[[var_name]], fill = cluster)
  ) +
    geom_bar(position = "fill") +           # proportion within each category
    coord_flip() +
    theme_minimal(base_size = 12) +
    labs(
      title = paste("Cluster Composition by", var_name),
      x = var_name,
      y = "Proportion"
    ) +
    theme(
      legend.position = "bottom",
      axis.text.y = element_text(size = 8)
    )
}

# By region
p_region <- plot_cluster_comp(kmodes_df_2, "Region")

# By season
p_season <- plot_cluster_comp(kmodes_df_2, "Season")

# By primary mode
p_primary <- plot_cluster_comp(kmodes_df_2, "Primary_mode")

# By etiology (or etiologic group, if you have that)
p_etiology <- plot_cluster_comp(kmodes_df_2, "Etiology_group")
```

Save plots

```{r}
# Create plots -----------------------------------------------------------

p_region <- plot_cluster_comp(kmodes_df_2, "Region")
p_season <- plot_cluster_comp(kmodes_df_2, "Season")
p_primary <- plot_cluster_comp(kmodes_df_2, "Primary_mode")
p_etiology <- plot_cluster_comp(kmodes_df_2, "Etiology_group")

# Save plots -------------------------------------------------------------

ggsave("cluster_region.png", p_region, width = 7, height = 5, dpi = 300)
ggsave("cluster_season.png", p_season, width = 7, height = 5, dpi = 300)
ggsave("cluster_primary_mode.png", p_primary, width = 7, height = 5, dpi = 300)
ggsave("cluster_etiology_group.png", p_etiology, width = 7, height = 5, dpi = 300)

```
