---
title: "DSE 511 Final Project Cluster Analysis in R"
author: "Cecilia Albert-Black"
date: "2025-11-11"
output: html_document
---

Set working directory to R code folder

```{r}
setwd("C:/Users/q4q/repos/DSE511_Final_Project/Step 3. Cluster Analysis/R Code")
```

Load necessary libraries (must be installed first)

```{r}
library(readr)
library(tidyverse)
library(fitdistrplus)
library(factoextra) #for pca
library(dbscan) # for pca
library(mclust)
library(klaR) # K-modes clustering
library(psych) # for KMO test
```

Import dataset from Step 1. Data Cleaning

```{r}
clean_df <- read_csv("C:/Users/q4q/repos/DSE511_Final_Project/Step 1. Data Cleaning/R code/outbreak_clean_r_v2.csv")

head(clean_df)
```

## Cluster Analysis

### I. Hierarchical

Encountered a limitation in R:

-   I tried running hierarchical clustering on all \~66,000 outbreaks

-   But hclust() in base R cannot handle more than 65,536 observations,

-   Even if it could, the distance matrix would be enormous (≈ 66k × 66k).

Thus, hierarchical clustering on a full distance matrix just doesn’t scale to this dataset.

### II. K-Means

Select variables to cluster

```{r}
num_vars <- clean_df %>%
  dplyr::select(
    Illnesses,
    Hospitalizations,
    Deaths,
    Hosp_per_ill,
    Death_per_ill)
```

Conduct a principal component analysis (PCA) to determine the number of clusters

```{r}
pca_result <- prcomp(num_vars, scale. = TRUE)
summary(pca_result)
plot(pca_result, type = "l")  # Scree plot
```

Extracted PCA scores

```{r}
pca_scores <- as.data.frame(pca_result$x[, 1:3])
head(pca_scores)
```

Run K-Means and set the seed to ensure reproducible cluster assignment

```{r}
set.seed(123)
kmeans_result <- kmeans(pca_scores, centers = 3, nstart = 25)
clean_df$cluster <- factor(kmeans_result$cluster)
```

Visualize the clusters on the PCA plot

```{r}
pca_clust <- fviz_cluster(kmeans_result, data = pca_scores, geom = "point", ellipse = TRUE)
pca_clust

# Save the plot
ggsave(
  filename = "PCA_kmeans_clusters.png",  # File name and extension
  plot = pca_clust,                      # Plot object to save
  width = 8,                             # Width in inches
  height = 6,                            # Height in inches
  dpi = 300                              # Resolution (good for publication)
)
```

It seems like there is one major outlier. My guess is that it's the 400,000 illnesses in March 1993 from a water borne outbreak we identified during the EDA.

### III. K-Modes

```{r}

```

### IV. K-Prototypes

```{r}

```

ChatGPT's comparison of clustering methods

| Method | Type | Strength | Limitation | Scales to 66k? |
|---------------|---------------|---------------|---------------|---------------|
| **K-means** | Partitioning | Fast, interpretable | Sensitive to outliers, assumes spherical clusters | ✅ |
| **Hierarchical** | Agglomerative | Visual dendrogram | Memory limit (\>65k rows) | ⚠️ (sample only) |
| **DBSCAN** | Density-based | Finds anomalies | Needs parameter tuning | ✅ (with PCA) |
| **Mclust (GMM)** | Model-based | Automatic k via BIC | Sensitive to outliers | ✅ |

## Cluster Characterization

#### a. By Numeric Variables

Examine numeric variable summaries

```{r}
clean_df %>%
  group_by(cluster) %>%
  summarise(across(
    c(Illnesses_std, Hospitalizations_std, Deaths_std, Hosp_per_ill, Death_per_ill),
    mean, na.rm = TRUE
  ))
```

It appears the following clusters could be described as follows

-   Cluster 1: severe outbreaks

-   Cluster 2: mild outbreaks

-   Cluster 3: anomaly outbreaks

Now lets compare the outbreaks by categorical variables

#### b. By Most Prevalent Primary Mode

```{r}
# 1. Identify top primary modes causing illnesses
top_mode <- clean_df %>%
  group_by(Primary_mode) %>%
  summarise(total_illnesses = sum(Illnesses, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_illnesses)) %>%
  pull(Primary_mode)   # <- this is the key fix

# 2. Filter dataset to those Primary_mode
df_top <- clean_df %>%
  filter(Primary_mode %in% top_mode)

# 3. Reorder Primary_mode factor by frequency
df_top <- df_top %>%
  mutate(Primary_mode = forcats::fct_infreq(Primary_mode))

# 4. Plot
ggplot(df_top, aes(x = Primary_mode, fill = cluster)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(
    title = "Cluster Composition by Primary Mode",
    y = "Proportion",
    x = "Primary_mode"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 8)
  )

# Save the plot
ggsave(
  filename = "Cluster_by_primary_mode.png",  # File name and extension
  width = 8,                             # Width in inches
  height = 6,                            # Height in inches
  dpi = 300                              # Resolution (good for publication)
)
```

As a table

```{r}
mode_cluster_tab <- table(clean_df$Primary_mode, clean_df$cluster)
mode_cluster_tab
```

It appears the following clusters could be described as follows

-   Cluster 1: severe outbreaks primarily caused by water borne illness

-   Cluster 2: mild outbreaks caused by many modes

-   Cluster 3: anomaly outbreaks caused by a water borne illness

#### c. By Most Prevalent Etiologies

```{r}
# 1. Identify top 50 etiologies causing illnesses
top_etiologies <- clean_df %>%
  group_by(Etiology) %>%
  summarise(total_illnesses = sum(Illnesses, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_illnesses)) %>%
  slice_head(n = 50) %>%
  pull(Etiology)   # <- this is the key fix

# 2. Filter dataset to those etiologies
df_top <- clean_df %>%
  filter(Etiology %in% top_etiologies)

# 3. Reorder Etiology factor by frequency
df_top <- df_top %>%
  mutate(Etiology = forcats::fct_infreq(Etiology))

# 4. Plot
ggplot(df_top, aes(x = Etiology, fill = cluster)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(
    title = "Cluster Composition by Top 50 Etiologies",
    y = "Proportion",
    x = "Etiology"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 8)
  )

# Save the plot
ggsave(
  filename = "Cluster_by_etiology.png",  # File name and extension
  width = 8,                             # Width in inches
  height = 6,                            # Height in inches
  dpi = 300                              # Resolution (good for publication)
)
```

It appears the following clusters could be further described as follows

-   Cluster 1: severe outbreaks primarily caused by water borne Listeria and Legionella

-   Cluster 2: mild outbreaks caused by many modes with no specific etiology

-   Cluster 3: anomaly outbreaks caused by caused by a water borne Cryptosporidium

#### d. By Most Prevalent Settings

```{r}
# 1. Identify top 50 settings causing illnesses
top_settings <- clean_df %>%
  group_by(Setting) %>%
  summarise(total_illnesses = sum(Illnesses, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_illnesses)) %>%
  slice_head(n = 50) %>%
  pull(Setting)   # <- this is the key fix

# 2. Filter dataset to those settings
df_top <- clean_df %>%
  filter(Setting %in% top_settings)

# 3. Reorder Setting factor by frequency
df_top <- df_top %>%
  mutate(Setting = forcats::fct_infreq(Setting))

# 4. Plot
ggplot(df_top, aes(x = Setting, fill = cluster)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(
    title = "Cluster Composition by Top 50 Settings",
    y = "Proportion",
    x = "Setting"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 8)
  )

# Save the plot
ggsave(
  filename = "Cluster_by_setting.png",  # File name and extension
  width = 8,                             # Width in inches
  height = 6,                            # Height in inches
  dpi = 300                              # Resolution (good for publication)
)
```

It appears the following clusters could be further described as follows

-   Cluster 1: severe outbreaks primarily caused by water borne Listeria and Legionella, spread primarily in Factories/Industrial Facilities or Hospitals

-   Cluster 2: mild outbreaks contracted in a variety of modes, with no specific etiology or setting

-   Cluster 3: anomaly outbreaks caused by caused by a water borne Cryptosporidium, spread in a community/municipality

#### e. By Season

```{r}
# Plot
ggplot(df_top, aes(x = Season, fill = cluster)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(
    title = "Cluster Composition by Season",
    y = "Proportion",
    x = "Season"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 8)
  )

# As a table
mode_cluster_tab <- table(clean_df$Season, clean_df$cluster)
mode_cluster_tab
```

It appears the following clusters could be further described as follows

-   Cluster 1: severe outbreaks primarily caused by water borne Listeria and Legionella, spread primarily in Factories/Industrial Facilities or Hospitals

-   Cluster 2: mild outbreaks contracted in a variety of modes, with no specific etiology or setting

-   Cluster 3: an anomaly outbreak caused by a water borne Cryptosporidium, spread in a community/municipality in the spring

#### f. By Year

```{r}
# Plot
ggplot(df_top, aes(x = Year, fill = cluster)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(
    title = "Cluster Composition by Year",
    y = "Proportion",
    x = "Season"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 8)
  )

# As a table
mode_cluster_tab <- table(clean_df$Year, clean_df$cluster)
mode_cluster_tab
```

It appears the following clusters could be further described as follows

-   Cluster 1: more severe outbreaks primarily caused by water borne illnesses like Listeria and Legionella, spread primarily in Factories/Industrial Facilities or Hospitals, occurring year-round but more prevalent prior to year 2000.

-   Cluster 2: mild outbreaks contracted in a variety of modes and settings, with no specific etiology or season.

-   Cluster 3: an anomaly outbreak caused by a water borne Cryptosporidium, spread in a community/municipality in the spring of 1993.

This validates our observation of the outlier value we observed in the EDA as being represented solely by cluster 3. This value was such an anomaly that it clustered as an individual point by itself.
