---
title: "DSE 511 Final Project: Data Cleaning"
author: "Cecilia Albert-Black"
date: "2025-11-04"
output: html_document
---

Set working directory to R code folder

```{r}
setwd("C:/Users/q4q/repos/DSE511_Final_Project/R code")
```

Load necessary libraries

```{r}
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(lubridate)
```

## Cleaning the NORS Outbreak Dataset

Import raw dataset from Raw data folder

```{r}
# Import
outbreak_df <- read_csv("C:/Users/q4q/repos/DSE511_Final_Project/Raw data/NORS_outbreaks.csv")

# Preview
head(outbreak_df)
```

### 1. Understanding/Inspecting the Dataset

Identify columns and their structure

```{r}
# Structure
str(outbreak_df)
```

Summary statistics by column using skimr package

```{r}
library(skimr)
skim(outbreak_df)
```

#### A. Categorical Variables (n = 12)

Determine unique values of categorical variables

##### I. States

The state where the exposure occurred. For a single state of exposure, the state will be listed. For multiple states of exposure, "Multistate" will be listed.

```{r}
unique(outbreak_df$State)
```

##### II. Primary Mode

Primary mode of transmission.

```{r}
# Table of unique primary modes
table(outbreak_df$`Primary Mode`)
```

##### III. Etiology

Genus and species of identified etiology. Multiple reported etiologies are separated by a semicolon.

```{r}
# Number of unique etiologies
n_distinct(outbreak_df$Etiology) 
```

```{r}
# Top 10 Etiologies
outbreak_df %>%
  count(Etiology, sort = TRUE)%>%
  head(20)
```

##### IV. Serotype or Genotype

Serotype or genotype of identified etiology. Multiple reported etiologies are separated by a semicolon.

```{r}
# Number of unique variables
n_distinct(outbreak_df$`Serotype or Genotype`)
```

Might be a difficult variable to handle since many are a list of descriptors; will be hard to categorize.

##### V. Etiology Status

Indicates whether or not the identified etiology is laboratory 'Confirmed' or 'Suspected'. Multiple reported etiologies are separated by a semicolon.

```{r}
# Number of unique variables
n_unique(outbreak_df$`Etiology Status`)
```

```{r}
# Top 10 unique names
unique(outbreak_df$`Etiology Status`)%>%
  head(10)   # top 10
```

Very messy variable that is hard to interpret - maybe we'll omit it or find a source that describes it better.

##### VI. Setting

Setting(s) where exposure occurred. For foodborne outbreaks, the setting used is the location where the food was prepared.

```{r}
n_unique(outbreak_df$Setting)
```

Many different settings, some have similarities, might be worth exploring ways to group them using cluster analysis.

##### VII. Food Vehicle

For foodborne outbreaks only, the implicated food. Multiple implicated foods are separated by a semicolon.

```{r}
# Top 10 food vehicles by rank
outbreak_df %>%
  count(`Food Vehicle`, sort = TRUE) %>%
  head(10)  
```

Investigate where the missing values of food vehicle are coming from

```{r}
outbreak_df %>%
  mutate(
    FoodVehicle_missing = if_else(
      is.na(`Food Vehicle`) | `Food Vehicle` == "", "Missing", "Present")) %>%
  count(`Primary Mode`, FoodVehicle_missing) %>%
  group_by(`Primary Mode`) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()

```

This table shows that about half of the reported food vehicles are missing within the category food within primary mode.

##### VIII. Food Contaminated Ingredient

For foodborne outbreaks only, indicates the contaminated ingredient. Multiple contaminated ingredients are separated by a semicolon.

```{r}
# Number of unique variables 
n_unique(outbreak_df$`Food Contaminated Ingredient`)
```

This might be a redundant variable, consider omitting and just using food vehicle for sake of simplicity.

##### IX. IFSAC Category

For foodborne outbreaks only, the IFSAC food category of the contaminated ingredient. If contaminated ingredients fall into multiple IFSAC categories, "Multiple" will be listed.

```{r}
unique(outbreak_df$`IFSAC Category`)
```

Confirm that this category only exists within Food Vehicle outbreaks and count how many missing values

```{r}
outbreak_df %>%
  group_by(`Primary Mode`) %>%
  summarize(
    total = n(),
    missing_IFSAC = sum(is.na(`IFSAC Category`)),
    percent_missing = 100 * missing_IFSAC / total
  ) %>%
  arrange(desc(percent_missing))

```

Looks good, about half of food borne outbreaks contain IFSAC Categories. Might not be much more informative than Food Vehicle for the sake of our study.

##### X. Water Exposure

For waterborne outbreaks only, the implicated type of water exposure.

```{r}
# Top 10 water exposure types by rank
outbreak_df %>%
  count(`Water Exposure`, sort = TRUE) %>%
  head(10)
```

Investigate where the missing values are coming from

```{r}
outbreak_df %>%
  mutate(
    Waterexposure_missing = if_else(
      is.na(`Water Exposure`) | `Water Exposure` == "", "Missing", "Present")) %>%
  count(`Primary Mode`, Waterexposure_missing) %>%
  group_by(`Primary Mode`) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()

```

All outbreaks within water as primary mode of exposure have a designated type of water exposure (in other words, no missing values of water exposure in water-borne outbreaks)

##### XI. Water Type

For waterborne outbreaks only, a description of the venue (for treated and untreated recreational water), water system (for drinking water) or device/structure (for other/unknown; e.g., steam cleaner, cooling tower, ornamental fountain, etc.) that was the vehicle for waterborne exposure to microbial pathogens, chemicals, or toxins.

```{r}
n_unique(outbreak_df$`Water Type`) 
```

```{r}
# Top 10 water types by rank
outbreak_df %>%
  count(`Water Type`, sort = TRUE)%>%
  head(10)
```

##### XII. Animal Type

For animal contact outbreaks only, the type of animal involved. Multiple animal types are separated by a semicolon.

```{r}
# Top 10 animal types by rank
outbreak_df %>%
  count(`Animal Type`, sort = TRUE) %>%
  head(10)
```

#### B. Numeric Variables (n = 7)

##### I. Year

Year of earliest date of reported illness onset.

```{r}
# Check if any years are missing
sum(is.na(outbreak_df$Year))
```

```{r}
# Count of outbreaks per year


# Visualize number of outbreaks per year
barplot(table(outbreak_df$Year),
        main = "Number of Outbreaks per Year",
        xlab = "Year",
        ylab = "Count of Outbreaks",
        col = "steelblue",
        border = NA)
```

Prior to 1998, outbreaks reported were \<1,000 maybe due to lack of proper reporting. Consider using data from 1998-present.

##### II. Month

Month of earliest date of reported illness onset.

```{r}
# Check if any months are missing within each year

missing_months_by_year <- outbreak_df %>%
  filter(!is.na(Year)) %>%
  group_by(Year) %>%
  summarize(
    observed_months = sort(unique(Month)),
    missing_months = list(setdiff(1:12, unique(Month))),
    n_missing = length(setdiff(1:12, unique(Month)))
  ) %>%
  filter(n_missing > 0)

missing_months_by_year

```

```{r}
# Visualize this using a heatmap

ggplot(outbreak_df, aes(x = Month, y = factor(Year))) +
  geom_tile(aes(fill = !is.na(Month)), color = "white") +
  scale_fill_manual(values = c("FALSE" = "grey90", "TRUE" = "steelblue")) +
  labs(title = "Month Coverage by Year",
       x = "Month",
       y = "Year",
       fill = "Data Present") +
  theme_minimal()

```

Prior to 1998, some months had no reported outbreaks. This could be true or also indicative of a weak reporting system.

```{r}
# Frequency of outbreaks by month across all years

month_freq <- outbreak_df %>%
  filter(!is.na(Month)) %>%
  count(Month, sort = TRUE)

month_freq
```

```{r}
# Visualization of frequency of outbreaks by month across all years
ggplot(outbreak_df, aes(x = factor(Month))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Frequency of Outbreaks per Month (All Years)",
       x = "Month",
       y = "Number of Outbreaks") +
  theme_minimal()
```

This shows the seasonal pattern in outbreak data; month may be a significant predictor of outbreak cases.

##### III. Illness

Estimated total number of primary cases, including lab-confirmed and probable, based on the outbreak-specific definition.

```{r}
# Histogram of Illnesses by outbreak
hist(outbreak_df$Illnesses)
```

Most outbreaks result in \<50 cases of illnesses.

```{r}
# Check for missing values
sum(is.na(outbreak_df$Illnesses))
```

##### IV. Hospitalizations

Number of primary cases who were hospitalized.

```{r}
hist(outbreak_df$Hospitalizations)
```

```{r}
# Check for missing values
sum(is.na(outbreak_df$Hospitalizations))
```

Several missing hospitalizations

##### V. Info on Hospitalizations

Total number of primary cases for whom information on hospitalization is available.

```{r}
unique(outbreak_df$`Info On Hospitalizations`)
```

This variable is hard to interpret and provides no additional information from the Hospitalizations variable.

##### VI. Deaths

Number of primary cases who died.

```{r}
hist(outbreak_df$Deaths)
```

```{r}
# Check for missing values
sum(is.na(outbreak_df$Deaths))
```

##### VII. Info on Deaths

Total number of primary cases for whom information on hospitalization is available.

```{r}
hist(outbreak_df$`Info On Deaths`)
```

### 2. Handling Missing Data

#### A. Removing Variables

-   Serotype or Genotype: many observations are a list of several types, might be hard to use this data.

-   Etiology Status: many observations are a list of several confirmed/suspected statuses by serotype or genotype, which would be hard to use.

-   Food Contaminated Ingredient: redundant variable, Food Type is enough.

-   Info on Hospitalizations: not actually indicative of hospitalizations.

-   Info on Deaths: not actually indicative of deaths.

```{r}
outbreak_df <- outbreak_df %>%
  select(
    -'Serotype or Genotype',
    -'Etiology Status',
    -'Info On Hospitalizations',
    -'Info On Deaths',
    -'Food Contaminated Ingredient'
  )

# Check
colnames(outbreak_df)
```

#### B. Impute NAs

-   Food Vehicle & IFSAC Category: NA values for Food Vehicle falling under the Food as Primary Mode of outbreak will be replaced by "Unknown".

-   Water Exposure & Water Type: leave NAs because this variable does not apply to non-waterborne outbreaks.

-   Animal Type: leave NAs because this variable does not apply to non-animal associated outbreaks.

-   Deaths & Hospitalizations: replace NAs with median value, to retain the same distribution of the variable while being robust to outliers (e.g. replacing with 0 or mean values would skew the data).

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    `Food Vehicle` = case_when(
      is.na(`Food Vehicle`) & 
        str_detect(`Primary Mode`, regex("Food", ignore_case = TRUE)) ~ "Unknown",
      TRUE ~ `Food Vehicle`
    ),
    Hospitalizations = if_else(
      is.na(Hospitalizations),
      median(Hospitalizations, na.rm = TRUE),
      Hospitalizations
    ),
    Deaths = if_else(
      is.na(Deaths),
      median(Deaths, na.rm = TRUE),
      Deaths
    )
  )
```

```{r}
# Check that no NAs exist in Food Vehicle when Food as Primary Mode of outbreak
outbreak_df %>%
  filter(str_detect(`Primary Mode`, regex("food", ignore_case = TRUE))) %>%
  summarize(
    missing_food_vehicle = sum(is.na(`Food Vehicle`))
  )

# Check that Unknown was successfully added as a new category of Food Vehicle for the NA values 
outbreak_df %>%
  count(`Food Vehicle`, sort = TRUE) %>%
  head(10)  
```

```{r}
# Check that NAs don't exist in hospitalizations and deaths anymore
outbreak_df %>%
  summarize(
    missing_hosp = sum(is.na(Hospitalizations)),
    missing_deaths = sum(is.na(Deaths))
  )
```

### 3. Standardize and Normalize Data

#### A. Standardize categorical variables (e.g., state abbreviations, illness names).

Create a date variable - to combine 2 time variables into 1

```{r}
# Create a Date variable which combines Month-Year
outbreak_df <- outbreak_df %>%
  mutate(
    Date = sprintf("%04d-%02d", Year, Month)  # formats as "YYYY-MM"
  )
```

Create a state abbreviation variable - for easier interpretation

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    state_abbr = state.abb[ match(State, state.name) ]
  )
```

#### B. Normalize numeric variables if you plan to use clustering or distance-based methods.

Standardize count data (create new variables)

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    Hospitalizations_std = scale(Hospitalizations),
    Deaths_std = scale(Deaths),
    Illnesses_std = scale(Illnesses)
  )

# Preview
outbreak_df %>%
  select(Hospitalizations_std, Deaths_std, Illnesses_std) %>%
  head(10)
```

### 4. Feature Engineering

#### I. Animal Group

```{r}
outbreak_df <- outbreak_df %>%
  mutate(Animal_group = case_when(
    is.na(`Animal Type`) ~ "Missing/NA",
    str_detect(`Animal Type`, ";") ~ "Multiple",  # catches multi-animal entries
    str_detect(`Animal Type`, regex("poultry", ignore_case = TRUE)) ~ "Poultry",
    str_detect(`Animal Type`, regex("cattle", ignore_case = TRUE)) ~ "Cattle",
    str_detect(`Animal Type`, regex("cat|dog|ferret|guinea pig|hedgehog|mouse|rat|raccoon", ignore_case = TRUE)) ~ "House Pets",
    str_detect(`Animal Type`, regex("horse|pony|goat|sheep|pig|alpaca|donkey|llama|yak", ignore_case = TRUE)) ~ "Other Ungulates",
    str_detect(`Animal Type`, regex("lizard|turtle|snake|reptile", ignore_case = TRUE)) ~ "Reptiles",
    str_detect(`Animal Type`, regex("bird, not including poultry", ignore_case = TRUE)) ~ "Other Birds",
    str_detect(`Animal Type`, regex("fish", ignore_case = TRUE)) ~ "Fish",
    str_detect(`Animal Type`, regex("frog|salamander", ignore_case = TRUE)) ~ "Amphibians",
    str_detect(`Animal Type`, regex("other", ignore_case = TRUE)) ~ "Other/Unknown",
    TRUE ~ "Other/Unknown"
  ))

# Check the counts for each group
outbreak_df %>%
  count(Animal_group, sort = TRUE)
```

#### II. Severity of Outbreak

Since these following variables are nested within each other illnesses(hospitalizations((deaths))), we'll create 3 new ratios to characterize outbreak severity.

```{r}
# Compute ratios only when denominator is > 0
outbreak_df <- outbreak_df %>%
  mutate(
    hosp_per_ill = if_else(Illnesses > 0, Hospitalizations / Illnesses, NA_real_),
    death_per_hosp = if_else(Hospitalizations > 0, Deaths / Hospitalizations, NA_real_),
    death_per_ill = if_else(Illnesses > 0, Deaths / Illnesses, NA_real_)
  )

# Mean ratios
outbreak_df %>%
  summarize(
    mean_hosp_per_ill = mean(hosp_per_ill, na.rm = TRUE),
    mean_death_per_hosp = mean(death_per_hosp, na.rm = TRUE),
    mean_death_per_ill = mean(death_per_ill, na.rm = TRUE)
  )
```

Check NAs of new ratio variables

```{r}
outbreak_df %>%
  summarize(
    na_hosp_per_ill = sum(is.na(hosp_per_ill)),
    na_death_per_hosp = sum(is.na(death_per_hosp)),
    na_death_per_ill = sum(is.na(death_per_ill))
  )
```

We'll omit death/hospitalizations since it will be hard to handle NAs

```{r}
outbreak_df <- outbreak_df %>%
  select(-death_per_hosp)
```

#### III. Binary Features for Primary Mode

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    is_food_related = if_else(str_detect(`Primary Mode`, regex("Food", ignore_case = TRUE)), 1, 0),
    is_water_related = if_else(str_detect(`Primary Mode`, regex("Water", ignore_case = TRUE)), 1, 0),
    is_animal_related = if_else(str_detect(`Primary Mode`, regex("Animal", ignore_case = TRUE)), 1, 0)
  )

# Preview the first few rows
outbreak_df %>%
  select(`Primary Mode`, is_food_related, is_water_related, is_animal_related) %>%
  head(10)
```

#### IV. Outbreak Month/Season

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5)  ~ "Spring",
      Month %in% c(6, 7, 8)  ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Fall",
      TRUE ~ NA_character_  # in case Month is missing or invalid
    )
  )

# Preview the first few rows
outbreak_df %>%
  select(Month, season) %>%
  head(10)
```

#### V. Reordering Columns

For readability

```{r}
# Define the desired column order
desired_order <- c(
  "Date", "Year", "Month", "season", "State", "state_abbr",
  "Primary Mode", "Etiology", "Serotype or Genotype", "IFSAC Category", "setting",
  "Illnesses", "Hospitalizations", "Deaths", "outbreak_severity",
  "food_vehicle_missing", "Water Exposure", "water_exposure_missing", "Water Type", "water_type_group",
  "Animal Type", "animal_group",
  "is_food_related", "is_water_related", "is_animal_related"
)

# Reorder columns, keeping only those that exist
outbreak_df <- outbreak_df %>%
  select(any_of(desired_order))

# Preview first 10 rows
outbreak_df %>%
  head(10)
```

### 5. Save to CSV

```{r}
# Save to CSV
write.csv(outbreak_df, "outbreak_clean_v3_r.csv", row.names = FALSE)
```
