---
title: "DSE 511 Final Project: Data Cleaning"
author: "Cecilia Albert-Black"
date: "2025-11-04"
output: html_document
---

Set working directory to R code folder

```{r}
setwd("C:/Users/q4q/repos/DSE511_Final_Project/Step 1. Data Cleaning/R code")
```

Load necessary libraries

```{r}
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(lubridate)
```

## Cleaning the NORS Outbreak Dataset

Import raw dataset from Raw data folder

```{r}
# Import
outbreak_df <- read_csv("C:/Users/q4q/repos/DSE511_Final_Project/Raw data/NORS_outbreaks.csv")

# Preview
head(outbreak_df)
```

### 1. Understanding/Inspecting the Dataset

Identify columns and their structure

```{r}
# Structure
str(outbreak_df)
```

Summary statistics by column using skimr package

```{r}
library(skimr)
skim(outbreak_df)
```

#### A. Categorical Variables (n = 12)

Determine unique values of categorical variables

##### I. States

The state where the exposure occurred. For a single state of exposure, the state will be listed. For multiple states of exposure, "Multistate" will be listed.

```{r}
unique(outbreak_df$State)
```

##### II. Primary Mode

Primary mode of transmission.

```{r}
# Table of unique primary modes
table(outbreak_df$`Primary Mode`)
```

Change variable name to one word

```{r}
outbreak_df$Primary_mode <- outbreak_df$'Primary Mode'
```

##### III. Etiology

Genus and species of identified etiology. Multiple reported etiologies are separated by a semicolon.

```{r}
# Number of unique etiologies
n_distinct(outbreak_df$Etiology) 
```

```{r}
# Top 10 Etiologies
outbreak_df %>%
  count(Etiology, sort = TRUE)%>%
  head(20)
```

##### IV. Serotype or Genotype

Serotype or genotype of identified etiology. Multiple reported etiologies are separated by a semicolon.

```{r}
# Number of unique variables
n_distinct(outbreak_df$`Serotype or Genotype`)
```

Might be a difficult variable to handle since many are a list of descriptors; will be hard to categorize.

##### V. Etiology Status

Indicates whether or not the identified etiology is laboratory 'Confirmed' or 'Suspected'. Multiple reported etiologies are separated by a semicolon.

```{r}
# Number of unique variables
n_unique(outbreak_df$`Etiology Status`)
```

```{r}
# Top 10 unique names
unique(outbreak_df$`Etiology Status`)%>%
  head(10)   # top 10
```

Very messy variable that is hard to interpret - maybe we'll omit it or find a source that describes it better.

##### VI. Setting

Setting(s) where exposure occurred. For foodborne outbreaks, the setting used is the location where the food was prepared.

```{r}
n_unique(outbreak_df$Setting)
```

Many different settings, some have similarities, might be worth exploring ways to group them using cluster analysis.

##### VII. Food Vehicle

For foodborne outbreaks only, the implicated food. Multiple implicated foods are separated by a semicolon.

```{r}
# Top 10 food vehicles by rank
outbreak_df %>%
  count(`Food Vehicle`, sort = TRUE) %>%
  head(10)  
```

Investigate where the missing values of food vehicle are coming from

```{r}
outbreak_df %>%
  mutate(
    FoodVehicle_missing = if_else(
      is.na(`Food Vehicle`) | `Food Vehicle` == "", "Missing", "Present")) %>%
  count(`Primary Mode`, FoodVehicle_missing) %>%
  group_by(`Primary Mode`) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()

```

This table shows that about half of the reported food vehicles are missing within the category food within primary mode.

Create a variable flagging missing food vehicle values

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    Food_vehicle_missing = if_else(
      is.na(`Food Vehicle`) | trimws(`Food Vehicle`) == "",
      "Missing",
      "Present"
    )
  )
```

Change variable name to one word

```{r}
outbreak_df$Food_vehicle <- outbreak_df$'Food Vehicle'
```

##### VIII. Food Contaminated Ingredient

For foodborne outbreaks only, indicates the contaminated ingredient. Multiple contaminated ingredients are separated by a semicolon.

```{r}
# Number of unique variables 
n_unique(outbreak_df$`Food Contaminated Ingredient`)
```

This might be a redundant variable, consider omitting and just using food vehicle for sake of simplicity.

##### IX. IFSAC Category

For foodborne outbreaks only, the IFSAC food category of the contaminated ingredient. If contaminated ingredients fall into multiple IFSAC categories, "Multiple" will be listed.

```{r}
unique(outbreak_df$`IFSAC Category`)
```

Confirm that this category only exists within Food Vehicle outbreaks and count how many missing values

```{r}
outbreak_df %>%
  group_by(`Primary Mode`) %>%
  summarize(
    total = n(),
    missing_IFSAC = sum(is.na(`IFSAC Category`)),
    percent_missing = 100 * missing_IFSAC / total
  ) %>%
  arrange(desc(percent_missing))

```

Looks good, about half of food borne outbreaks contain IFSAC Categories. Might not be much more informative than Food Vehicle for the sake of our study.

Change variable name to be one word

```{r}
outbreak_df$IFSAC_category <- outbreak_df$'IFSAC Category'
```

##### X. Water Exposure

For waterborne outbreaks only, the implicated type of water exposure.

```{r}
# Top 10 water exposure types by rank
outbreak_df %>%
  count(`Water Exposure`, sort = TRUE) %>%
  head(10)
```

Investigate where the missing values are coming from

```{r}
# Create a column flagging missing vs present values
outbreak_df$Water_exposure_missing <- ifelse(
  is.na(outbreak_df$`Water Exposure`) | trimws(outbreak_df$`Water Exposure`) == "",
  "Missing",
  "Present"
)

# Count by primary mode and missing/present status
outbreak_df %>%
  mutate(
    Waterexposure_missing = if_else(
      is.na(`Water Exposure`) | `Water Exposure` == "", "Missing", "Present")) %>%
  count(`Primary Mode`, Waterexposure_missing) %>%
  group_by(`Primary Mode`) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()
```

All outbreaks within water as primary mode of exposure have a designated type of water exposure (in other words, no missing values of water exposure in water-borne outbreaks).

Change variable name to be one word

```{r}
outbreak_df$Water_exposure <- outbreak_df$'Water Exposure'
```

##### XI. Water Type

For waterborne outbreaks only, a description of the venue (for treated and untreated recreational water), water system (for drinking water) or device/structure (for other/unknown; e.g., steam cleaner, cooling tower, ornamental fountain, etc.) that was the vehicle for waterborne exposure to microbial pathogens, chemicals, or toxins.

```{r}
n_unique(outbreak_df$`Water Type`) 
```

```{r}
# Top 10 water types by rank
outbreak_df %>%
  count(`Water Type`, sort = TRUE)%>%
  head(10)
```

Change variable name to be one word

```{r}
outbreak_df$Water_type <- outbreak_df$'Water Type'
```

##### XII. Animal Type

For animal contact outbreaks only, the type of animal involved. Multiple animal types are separated by a semicolon.

```{r}
# Top 10 animal types by rank
outbreak_df %>%
  count(`Animal Type`, sort = TRUE) %>%
  head(10)
```

Change variable name to be one word

```{r}
outbreak_df$Animal_type <- outbreak_df$'Animal Type'
```

#### B. Numeric Variables (n = 7)

##### I. Year

Year of earliest date of reported illness onset.

```{r}
# Check if any years are missing
sum(is.na(outbreak_df$Year))
```

```{r}
# Count of outbreaks per year


# Visualize number of outbreaks per year
barplot(table(outbreak_df$Year),
        main = "Number of Outbreaks per Year",
        xlab = "Year",
        ylab = "Count of Outbreaks",
        col = "steelblue",
        border = NA)
```

Prior to 1998, outbreaks reported were \<1,000 maybe due to lack of proper reporting. Consider using data from 1998-present.

##### II. Month

Month of earliest date of reported illness onset.

```{r}
# Check if any months are missing within each year

missing_months_by_year <- outbreak_df %>%
  filter(!is.na(Year)) %>%
  group_by(Year) %>%
  summarize(
    observed_months = sort(unique(Month)),
    missing_months = list(setdiff(1:12, unique(Month))),
    n_missing = length(setdiff(1:12, unique(Month)))
  ) %>%
  filter(n_missing > 0)

missing_months_by_year

```

```{r}
# Visualize this using a heatmap

ggplot(outbreak_df, aes(x = Month, y = factor(Year))) +
  geom_tile(aes(fill = !is.na(Month)), color = "white") +
  scale_fill_manual(values = c("FALSE" = "grey90", "TRUE" = "steelblue")) +
  labs(title = "Month Coverage by Year",
       x = "Month",
       y = "Year",
       fill = "Data Present") +
  theme_minimal()

```

Prior to 1998, some months had no reported outbreaks. This could be true or also indicative of a weak reporting system.

```{r}
# Frequency of outbreaks by month across all years

month_freq <- outbreak_df %>%
  filter(!is.na(Month)) %>%
  count(Month, sort = TRUE)

month_freq
```

```{r}
# Visualization of frequency of outbreaks by month across all years
ggplot(outbreak_df, aes(x = factor(Month))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Frequency of Outbreaks per Month (All Years)",
       x = "Month",
       y = "Number of Outbreaks") +
  theme_minimal()
```

This shows the seasonal pattern in outbreak data; month may be a significant predictor of outbreak cases.

##### III. Illness

Estimated total number of primary cases, including lab-confirmed and probable, based on the outbreak-specific definition.

```{r}
# Histogram of Illnesses by outbreak
hist(outbreak_df$Illnesses)
```

Most outbreaks result in \<50 cases of illnesses.

```{r}
# Check for missing values
sum(is.na(outbreak_df$Illnesses))
```

##### IV. Hospitalizations

Number of primary cases who were hospitalized.

```{r}
hist(outbreak_df$Hospitalizations)
```

```{r}
# Check for missing values
sum(is.na(outbreak_df$Hospitalizations))
```

Several missing hospitalizations

##### V. Info on Hospitalizations

Total number of primary cases for whom information on hospitalization is available.

```{r}
unique(outbreak_df$`Info On Hospitalizations`)
```

This variable is hard to interpret and provides no additional information from the Hospitalizations variable.

##### VI. Deaths

Number of primary cases who died.

```{r}
hist(outbreak_df$Deaths)
```

```{r}
# Check for missing values
sum(is.na(outbreak_df$Deaths))
```

##### VII. Info on Deaths

Total number of primary cases for whom information on hospitalization is available.

```{r}
hist(outbreak_df$`Info On Deaths`)
```

### 2. Handling Missing Data

Remove columns with mostly missing values

```{r}
# Call the columns to drop
columns_to_drop <- c(
  "Serotype or Genotype",
  "Info On Hospitalizations",
  "Info On Deaths",
  "Etiology Status",
  "Food Contaminated Ingredient"
)

# Drop columns
outbreak_df <- outbreak_df %>%
  dplyr::select(-all_of(columns_to_drop))

# Check
names(outbreak_df)
```

#### B. Impute NAs

```{r}
# ---- Define column groups ----
categorical_cols <- c("Etiology", "Setting")

contingent_cols <- c(
  "Food Vehicle",
  "IFSAC Category",
  "Water Exposure",
  "Water Type",
  "Animal Type"
)

numeric_cols <- c("Hospitalizations", "Deaths")

# ---- Fill missing values ----
outbreak_df <- outbreak_df %>%
  mutate(
    # Categorical columns → replace NA with "Unknown"
    across(all_of(categorical_cols), ~ replace_na(., "Unknown")),
    
    # Contingent categorical columns → replace NA with "Not Applicable"
    across(all_of(contingent_cols), ~ replace_na(., "Not Applicable")),
    
    # Numeric columns → replace NA with median
    across(all_of(numeric_cols), ~ if_else(is.na(.), median(., na.rm = TRUE), .))
  )

# ---- Check missing values after cleaning ----
missing_summary <- outbreak_df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
  arrange(desc(missing_count))

print("Missing values after cleaning:")
print(missing_summary, n = Inf)
```

### 3. Standardize and Normalize Data

Standardize count data (create new variables)

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    Hospitalizations_std = scale(Hospitalizations),
    Deaths_std = scale(Deaths),
    Illnesses_std = scale(Illnesses)
  )

# Preview
outbreak_df %>%
  dplyr::select(Hospitalizations_std, Deaths_std, Illnesses_std) %>%
  head(10)
```

### 4. Feature Engineering

#### I. Animal Group

```{r}
outbreak_df <- outbreak_df %>%
  mutate(Animal_type_group = case_when(
    is.na(`Animal Type`) ~ "Missing/NA",
    str_detect(`Animal Type`, ";") ~ "Multiple",  # catches multi-animal entries
    str_detect(`Animal Type`, regex("poultry", ignore_case = TRUE)) ~ "Poultry",
    str_detect(`Animal Type`, regex("cattle", ignore_case = TRUE)) ~ "Cattle",
    str_detect(`Animal Type`, regex("cat|dog|ferret|guinea pig|hedgehog|mouse|rat|raccoon", ignore_case = TRUE)) ~ "House Pets",
    str_detect(`Animal Type`, regex("horse|pony|goat|sheep|pig|alpaca|donkey|llama|yak", ignore_case = TRUE)) ~ "Other Ungulates",
    str_detect(`Animal Type`, regex("lizard|turtle|snake|reptile", ignore_case = TRUE)) ~ "Reptiles",
    str_detect(`Animal Type`, regex("bird, not including poultry", ignore_case = TRUE)) ~ "Other Birds",
    str_detect(`Animal Type`, regex("fish", ignore_case = TRUE)) ~ "Fish",
    str_detect(`Animal Type`, regex("frog|salamander", ignore_case = TRUE)) ~ "Amphibians",
    str_detect(`Animal Type`, regex("other", ignore_case = TRUE)) ~ "Other/Unknown",
    TRUE ~ "Other/Unknown"
  ))

# Check the counts for each group
outbreak_df %>%
  count(Animal_type_group, sort = TRUE)
```

#### II. Etiology Group

Create a variable that groups etiologies

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    Etiology_group = case_when(
      grepl("norovirus", Etiology, ignore.case = TRUE) ~ "Norovirus",
      grepl("salmonella", Etiology, ignore.case = TRUE) ~ "Salmonella",
      grepl("campylobacter", Etiology, ignore.case = TRUE) ~ "Campylobacter",
      grepl("clostridium", Etiology, ignore.case = TRUE) ~ "Clostridium",
      grepl("staphylococcus", Etiology, ignore.case = TRUE) ~ "Staphylococcus",
      grepl("hepatitis", Etiology, ignore.case = TRUE) ~ "Hepatitis",
      grepl("unknown|unspecified|not determined", Etiology, ignore.case = TRUE) ~ "Unknown",
      TRUE ~ "Other"
    )
  )

# Count values by group
outbreak_df %>%
  count(Etiology_group, sort = TRUE)
```

#### III. Region

```{r}
library(dplyr)

outbreak_df <- outbreak_df %>%
  mutate(
    Region = case_when(
      # Northeast
      State %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire",
                   "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
      
      # Midwest
      State %in% c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin",
                   "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska",
                   "North Dakota", "South Dakota") ~ "Midwest",
      
      # South
      State %in% c("Delaware", "District of Columbia", "Florida", "Georgia",
                   "Maryland", "North Carolina", "South Carolina", "Virginia",
                   "West Virginia",
                   "Alabama", "Kentucky", "Mississippi", "Tennessee",
                   "Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "South",
      
      # West
      State %in% c("Arizona", "Colorado", "Idaho", "Montana", "Nevada",
                   "New Mexico", "Utah", "Wyoming",
                   "Alaska", "California", "Hawaii", "Oregon", 
                   "Washington") ~ "West",
      
      # Territories
      State %in% c("Puerto Rico", "Guam", "Northern Mariana Islands",
                   "Virgin Islands", "Republic of Palau", 
                   "Republic of the Marshall Islands") ~ "Territory",
      
      # Multistate special case
      State == "Multistate" ~ "Multistate",
      
      TRUE ~ "Other"
    )
  )
```

#### IV. Water Type Group

Create a new variable which groups water types

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    Water_type_group = case_when(
      is.na(`Water Type`) ~ "Missing/NA",
      grepl("community|individual|private", `Water Type`, ignore.case = TRUE) ~ "Drinking/Community Water",
      grepl("pool|hot tub|spa|whirlpool|splash pad|fountain|water playground", `Water Type`, ignore.case = TRUE) ~ "Recreational Water",
      grepl("lake|reservoir|other", `Water Type`, ignore.case = TRUE) ~ "Natural/Other",
      grepl("unknown|unspecified|not determined", `Water Type`, ignore.case = TRUE) ~ "Unknown",
      TRUE ~ "Other"
    )
  )

# View counts of each group
outbreak_df %>%
  count(Water_type_group, sort = TRUE)
```

#### V. Setting Group

```{r}
library(dplyr)
library(stringr)

outbreak_df <- outbreak_df %>%
  mutate(
    Setting_group = case_when(
      
      # Food Service
      str_detect(Setting, regex("Restaurant|Caterer|Grocery|Store", ignore_case = TRUE)) ~ "Food Service",
      
      # Home/Residential
      str_detect(Setting, regex("Private home|Apartment|Mobile home|residence", ignore_case = TRUE)) ~ "Home/Residential",
      
      # Healthcare
      str_detect(Setting, regex("Hospital|nursing home|assisted living|long-term care", ignore_case = TRUE)) ~ "Healthcare",
      
      # Education
      str_detect(Setting, regex("School|college|university|daycare|preschool", ignore_case = TRUE)) ~ "Education",
      
      # Workplace/Industrial
      str_detect(Setting, regex("Factory|industrial|workplace|Place of Work", ignore_case = TRUE)) ~ "Workplace/Industrial",
      
      # Community/Public Facilities
      str_detect(Setting, regex("Community|municipality|Subdivision|Hall|Fee-based|Correctional|detention|Native American", ignore_case = TRUE)) ~ "Community/Public",
      
      # Recreation / Parks / Outdoors
      str_detect(Setting, regex("Park|Resort|Hotel|Beach|Camp|cabin|Backcountry|wilderness|Zoo|Festival|Fair", ignore_case = TRUE)) ~ "Recreation/Outdoors",
      
      # Unknown / Missing
      str_detect(Setting, regex("Unknown", ignore_case = TRUE)) ~ "Unknown",
      
      # Catch-all
      TRUE ~ "Other"
    )
  )
```

Check for NAs in new category variables (should be 0).

```{r}
library(dplyr)
outbreak_df %>%
  summarize(
    na_Animal_type_group = sum(is.na(Animal_type_group)),
    na_Etiology_group = sum(is.na(Etiology_group)),
    na_Region = sum(is.na(Region)),
    na_Water_type_group = sum(is.na(Water_type_group)),
    na_Setting_group = sum(is.na(Setting_group))
  )
```

#### VI. Severity of Outbreak

Since these following variables are nested within each other illnesses(hospitalizations((deaths))), we'll create 3 new ratios to characterize outbreak severity.

```{r}
# Compute ratios only when denominator is > 0
outbreak_df <- outbreak_df %>%
  mutate(
    Hosp_per_ill = if_else(Illnesses > 0, Hospitalizations / Illnesses, NA_real_),
    Death_per_hosp = if_else(Hospitalizations > 0, Deaths / Hospitalizations, NA_real_),
    Death_per_ill = if_else(Illnesses > 0, Deaths / Illnesses, NA_real_)
  )

# Mean ratios
outbreak_df %>%
  summarize(
    Mean_hosp_per_ill = mean(Hosp_per_ill, na.rm = TRUE),
    Mean_death_per_hosp = mean(Death_per_hosp, na.rm = TRUE),
    Mean_death_per_ill = mean(Death_per_ill, na.rm = TRUE)
  )
```

Check NAs of new ratio variables

```{r}
outbreak_df %>%
  summarize(
    na_hosp_per_ill = sum(is.na(Hosp_per_ill)),
    na_death_per_hosp = sum(is.na(Death_per_hosp)),
    na_death_per_ill = sum(is.na(Death_per_ill))
  )
```

We'll omit death/hospitalizations since it will be hard to handle NAs

```{r}
outbreak_df <- outbreak_df %>%
  dplyr::select(-Death_per_hosp)
```

#### V. Binary Features for Primary Mode

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    is_food_related = if_else(str_detect(`Primary Mode`, regex("Food", ignore_case = TRUE)), 1, 0),
    is_water_related = if_else(str_detect(`Primary Mode`, regex("Water", ignore_case = TRUE)), 1, 0),
    is_animal_related = if_else(str_detect(`Primary Mode`, regex("Animal", ignore_case = TRUE)), 1, 0)
  )

# Preview the first few rows
outbreak_df %>%
  dplyr::select(`Primary Mode`, is_food_related, is_water_related,
                is_animal_related) %>%
  head(10)
```

#### VI. Outbreak Month/Season

```{r}
outbreak_df <- outbreak_df %>%
  mutate(
    Season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5)  ~ "Spring",
      Month %in% c(6, 7, 8)  ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Fall",
      TRUE ~ NA_character_  # in case Month is missing or invalid
    )
  )

# Preview the first few rows
outbreak_df %>%
  dplyr::select(Month, Season) %>%
  head(10)
```

Create a date variable

```{r}
# Create a Date variable which combines Month-Year
outbreak_df <- outbreak_df %>%
  mutate(
    Date = sprintf("%04d-%02d", Year, Month)  # formats as "YYYY-MM"
  )
```

#### VII. Reordering Columns

For readability

```{r}
library(dplyr)

# Define the desired column order
desired_order <- c(
  "Year", "Month", "Date", "Season", "State", "Region",
  "Primary_mode", "Setting", "Setting_group", "Etiology", "Etiology_group",
  "IFSAC_category", "Food_vehicle", 
  "Water_exposure", "Water_type", "Water_type_group",
  "Animal_type", "Animal_type_group", 
  "Illnesses", "Hospitalizations", "Deaths",
  "Hospitalizations_std", "Deaths_std", "Illnesses_std", 
  "Hosp_per_ill", "Death_per_ill", 
  "Food_vehicle_missing", "Water_exposure_missing",
  "is_food_related", "is_water_related", "is_animal_related"
)

# Reorder columns, keeping only those that exist
outbreak_df <- outbreak_df %>%
  dplyr::select(any_of(desired_order))

# Preview first 10 rows
outbreak_df %>%
  head(10)
```

### 5. Save to CSV

```{r}
# Save to CSV
write.csv(outbreak_df, "outbreak_clean_r_v3.csv", row.names = FALSE)
```
