---
title: "DSE 511 Final Project: Exploratory Data Analysis"
author: "Cecilia Albert-Black"
date: "2025-11-07"
output: html_document
---

Set working directory to R code folder

```{r}
setwd("C:/Users/q4q/repos/DSE511_Final_Project/Step 2. Exploratory Data Analysis/R Code")
```

Load necessary libraries

```{r}
library(readr)
library(tidyverse)
library(skimr)
library(scales)
```

## Exploratory Data Analysis of Clean Dataset

Import dataset from Step 1. Data Cleaning

```{r}
clean_df <- read_csv("C:/Users/q4q/repos/DSE511_Final_Project/Step 1. Data Cleaning/R code/outbreak_clean_r_v2.csv")

head(clean_df)
```

## I. Basic Dataset Overview

Check structure, missing data, and basic info.

```{r}
str(clean_df)
```

Summary statistics

```{r}
skim(clean_df)
```

Check missing values again

```{r}
missing_summary <- data.frame(
  missing_values = sort(colSums(is.na(clean_df)), decreasing = TRUE)
)
missing_summary
```

## II. Numeric Variable Distributions

Explore the distributions of numeric variables

```{r}
# 1. Define numeric variables
num_vars <- c(
  "Illnesses", "Hospitalizations", "Deaths",
  "Illnesses_std", "Hospitalizations_std", "Deaths_std",
  "Hosp_per_ill", "Death_per_ill"
)

# 2. Reshape the numeric data for plotting
num_vars_long <- clean_df %>%
  select(all_of(num_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

# 3. Plot histograms for each numeric variable
ggplot(num_vars_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Histograms of Numeric Variables",
    x = "Value",
    y = "Frequency"
  )

# 4. Save to repo
ggsave(
  filename = "Histogram_numeric_vars.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

It appears that all variables will need to be log-transformed to achieve are more normal distribution

```{r}
# 1. Define variables to log-transform
vars_to_log <- c(
  "Illnesses", "Hospitalizations", "Deaths",
  "Illnesses_std", "Hospitalizations_std", "Deaths_std",
  "Hosp_per_ill", "Death_per_ill"
)

# 2. Apply log transformation (add small constant to handle zeros)
clean_df <- clean_df %>%
  mutate(
    across(
      all_of(vars_to_log),
      ~ log(. + 1),                # add 1 to avoid log(0)
      .names = "log_{.col}"        # create new columns prefixed with 'log_'
    )
  )

# 3. Reshape data for easy plotting
log_vars_long <- clean_df %>%
  select(starts_with("log_")) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

# 4. Plot histograms for each log variable and add a kernel density estimation
ggplot(log_vars_long, aes(x = value)) +
  # histogram as density
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "steelblue", color = "white", alpha = 0.6) +
  # kernel density estimate
  geom_density(linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Histograms and Kernel Density Estimates of Log-Transformed Variables",
    x = "Log-transformed value",
    y = "Density"
  )
# 5. Save to repo
ggsave(
  filename = "Histograms_and_KDE_of_log_vars.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

Count data might not be normalized, consider fitting other distribution functions.

## III. Top Contributing Factors

### Etiologies

#### a. Top 50 Etiologies Resulting in the Most Total Number of Illnesses

```{r}
# 1. Isolate top 50 etiologies by total illnesses
top_etiology <- clean_df %>%
  group_by(Etiology) %>%
  summarise(total_illnesses = sum(Illnesses, na.rm = TRUE)) %>%
  arrange(desc(total_illnesses)) %>%
  slice_head(n = 50)

# 2. Plot
ggplot(top_etiology,
       aes(y = reorder(Etiology, total_illnesses),
           x = total_illnesses,
           fill = Etiology)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "mako") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 50 Etiologies by Total Illnesses",
    x = "Total Illnesses",
    y = "Etiology"
  )

# 3. Save to repo
ggsave(
  filename = "Top_etiologies_by_total_illnesses.png",  # file name
  width = 12, height = 12, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

#### b. Top 50 Etiologies by Hospitalization Severity

```{r}
# 1. Calculate median severity ratios by etiology 
severity <- clean_df %>%
  group_by(Etiology) %>%
  summarise(
    median_hosp_per_ill = median(Hosp_per_ill, na.rm = TRUE)
  ) %>%
  arrange(desc(median_hosp_per_ill)) %>%
  slice_head(n = 50)  

# 2. Clean up long labels for plotting 
library(stringr)

severity <- severity %>%
  mutate(Etiology_lab = str_wrap(Etiology, width = 35))  # wrap long names

# 3. Make the plot object
ggplot(
  severity,
  aes(
    x = median_hosp_per_ill,
    y = reorder(Etiology_lab, median_hosp_per_ill),
    fill = Etiology
  )
) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "mako") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 50 Etiologies by Median Hospitalizations per Illness",
    x = "Median Hospitalizations per Illness",
    y = "Etiology"
  )


# 4. Save to file
ggsave(
  filename = "Top_etiologies_by_median_hospitalizations_per_illness.png", 
  width = 12, height = 12,
  units = "in",
  dpi = 300
)
```

Etiologies which have a hospitalization/illness rate \>50%

```{r}
# Identify etiologies where median Hosp_per_ill > 0.5 (i.e., >50%)
high_severity <- severity %>%
  filter(median_hosp_per_ill > 0.5)

# View results
high_severity

```

#### d. Top 50 Etiologies by Death Severity

```{r}
# 1. Calculate median severity ratios by etiology 
severity <- clean_df %>%
  group_by(Etiology) %>%
  summarise(
    median_death_per_ill = median(Death_per_ill, na.rm = TRUE)
  ) %>%
  arrange(desc(median_death_per_ill)) %>%
  slice_head(n = 50)  

# 2. Clean up long labels for plotting 
library(stringr)

severity <- severity %>%
  mutate(Etiology_lab = str_wrap(Etiology, width = 35))  # wrap long names

# 3. Make the plot object
ggplot(
  severity,
  aes(
    x = median_death_per_ill,
    y = reorder(Etiology_lab, median_death_per_ill),
    fill = Etiology
  )
) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "mako") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 50 Etiologies by Median Death per Illness",
    x = "Median Death per Illness",
    y = "Etiology"
  )


# 4. Save to file
ggsave(
  filename = "Top_etiologies_by_median_death_per_illness.png", 
  width = 12, height = 12,
  units = "in",
  dpi = 300
)
```

Etiologies with a high death/illness rate, say above 20%

```{r}
# Identify etiologies where median Death_per_ill > 0.5 (i.e., >20%)
high_lethality <- severity %>%
  filter(median_death_per_ill > 0.20)

# View results
high_lethality

```

### Settings

#### a. Top 50 Settings in Which Illnesses Occur

```{r}
# Isolate top 50 settings by total illnesses
top_setting <- clean_df %>%
  group_by(Setting) %>%
  summarise(total_illnesses = sum(Illnesses, na.rm = TRUE)) %>%
  arrange(desc(total_illnesses)) %>%
  slice_head(n = 50)

# 2. Plot
ggplot(top_setting,
       aes(y = reorder(Setting, total_illnesses),
           x = total_illnesses,
           fill = Setting)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "C") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 50 Settings by Total Illnesses",
    x = "Total Illnesses",
    y = "Setting"
  )

# 3. Save to repo
ggsave(
  filename = "Top_settings_by_total_illnesses.png",
  width = 12, height = 12, units = "in", dpi = 200
)
```

### Primary Mode

#### a. Most Risky Primary Mode of Outbreak

```{r}
primary_mode_summary <- clean_df %>%
  group_by(Primary_mode) %>%
  summarise(
    median_illnesses     = median(Illnesses, na.rm = TRUE),
    mean_hosp_per_ill    = mean(Hosp_per_ill, na.rm = TRUE),
    mean_death_per_ill   = mean(Death_per_ill, na.rm = TRUE),
    n_outbreaks          = n(),   # gives context about reporting volume
    .groups = "drop"
  ) %>%
  arrange(desc(median_illnesses))

primary_mode_summary
```

Plot of illnesses by primary mode

```{r}
# Plot median illnesses by primary mode
ggplot(primary_mode_summary, aes(x = reorder(Primary_mode, -median_illnesses), y = median_illnesses, fill = Primary_mode)) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Median Illnesses by Primary Mode",
    x = "Primary Mode",
    y = "Median Illnesses per Outbreak"
  )

# Save
ggsave("Median_Illnesses_by_Primary_Mode.png", width = 8, height = 5, units = "in", dpi = 300)
```

Plot of hospitalizations/illnesses by primary mode

```{r}
ggplot(primary_mode_summary, aes(x = reorder(Primary_mode, -median_illnesses), y = mean_hosp_per_ill, fill = Primary_mode)) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Mean Hosp/Ill by Primary Mode",
    x = "Primary Mode",
    y = "Mean Hosp/Ill per Outbreak"
  )

# Save
ggsave("Mean_Hosp_per_Ill_by_Primary_Mode.png", width = 8, height = 5, units = "in", dpi = 300)
```

Plot of deaths/illnesses by primary mode

```{r}
ggplot(primary_mode_summary, aes(x = reorder(Primary_mode, -median_illnesses), y = mean_death_per_ill, fill = Primary_mode)) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Mean Death/Ill by Primary Mode",
    x = "Primary Mode",
    y = "Mean Death/Ill per Outbreak"
  )

# Save
ggsave("Mean_Death_per_Ill_by_Primary_Mode.png", width = 8, height = 5, units = "in", dpi = 300)
```

## Spatial Trends

### a. Median Illness by State

```{r}
# 1. Compute median illnesses per state
geo <- clean_df %>%
  group_by(State_abbr) %>%
  summarise(median_illnesses = median(Illnesses, na.rm = TRUE)) %>%
  arrange(desc(median_illnesses)) %>%
  slice_head(n = 15)

# 2. Plot top 15 states by median illnesses
ggplot(
  geo,
  aes(
    x = reorder(State_abbr, -median_illnesses),
    y = median_illnesses,
    fill = State_abbr
  )
) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 15 States by Median Number of Illnesses",
    x = "State",
    y = "Median Number of Illnesses"
  )

# 3. Save the plot
ggsave(
  filename = "Top_15_States_by_Median_Illnesses.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

### b. Median Yearly Illness by State

```{r}
# 1. Compute total illnesses per state-year, then median yearly illness rate per state
yearly_geo <- clean_df %>%
  group_by(State_abbr, Year) %>%
  summarise(yearly_illnesses = sum(Illnesses, na.rm = TRUE), .groups = "drop") %>%
  group_by(State_abbr) %>%
  summarise(median_yearly_illnesses = median(yearly_illnesses, na.rm = TRUE)) %>%
  arrange(desc(median_yearly_illnesses)) %>%
  slice_head(n = 15)

# 2. Plot top 15 states by median yearly illness rate
ggplot(
  yearly_geo,
  aes(
    x = reorder(State_abbr, -median_yearly_illnesses),
    y = median_yearly_illnesses,
    fill = State_abbr
  )
) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 15 States by Median Yearly Illness Rate",
    x = "State",
    y = "Median Yearly Illnesses"
  )

# 3. Save the plot
ggsave(
  filename = "Top_15_States_by_Median_Yearly_Illnesses.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

### c. Mean Yearly Hospitalization/Illness Rate by State

```{r}
# 1. Compute mean Hosp_per_ill per state-year, then mean across years per state
yearly_hosp_ratio <- clean_df %>%
  group_by(State_abbr, Year) %>%
  summarise(yearly_hosp_ratio = mean(Hosp_per_ill, na.rm = TRUE), .groups = "drop") %>%
  group_by(State_abbr) %>%
  summarise(mean_yearly_hosp_ratio = mean(yearly_hosp_ratio, na.rm = TRUE)) %>%
  arrange(desc(mean_yearly_hosp_ratio)) %>%
  slice_head(n = 15)

# 2. Plot top 15 states by mean yearly hospitalization ratio
ggplot(
  yearly_hosp_ratio,
  aes(
    x = reorder(State_abbr, -mean_yearly_hosp_ratio),
    y = mean_yearly_hosp_ratio,
    fill = State_abbr
  )
) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 15 States by Mean Yearly Hospitalization/Illnesses",
    x = "State",
    y = "Mean Yearly Hospitalization Ratio"
  )

# 3. Save plot
ggsave(
  filename = "Top_15_States_by_Mean_Yearly_Hosp_per_Ill.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

### d. Mean Yearly Deaths/Illness Rate by State

```{r}
# 1. Compute mean Death_per_ill per state-year, then mean across years per state
yearly_death_ratio <- clean_df %>%
  group_by(State_abbr, Year) %>%
  summarise(yearly_death_ratio = mean(Death_per_ill, na.rm = TRUE), .groups = "drop") %>%
  group_by(State_abbr) %>%
  summarise(mean_yearly_death_ratio = mean(yearly_death_ratio, na.rm = TRUE)) %>%
  arrange(desc(mean_yearly_death_ratio)) %>%
  slice_head(n = 15)

# 2. Plot top 15 states by mean yearly death ratio
ggplot(
  yearly_death_ratio,
  aes(
    x = reorder(State_abbr, -mean_yearly_death_ratio),
    y = mean_yearly_death_ratio,
    fill = State_abbr
  )
) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 15 States by Mean Yearly Deaths/Illnesses",
    x = "State",
    y = "Mean Yearly Death Ratio"
  )

# 3. Save the plot
ggsave(
  filename = "Top_15_States_by_Mean_Yearly_Death_per_Ill.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

## Temporal Trends

### a. Seasonal Decomposition

Median illness by season across all types of outbreaks and dates

```{r}
# 1. Median illnesses by season
seasonal <- clean_df %>%
  group_by(Season) %>%
  summarise(median_illnesses = median(Illnesses, na.rm = TRUE)) %>%
  ungroup()

# 2. Plot
ggplot(seasonal, aes(x = Season, y = median_illnesses, fill = Season)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Median Illnesses by Season",
    x = "Season",
    y = "Median Illnesses"
  )

# 3. Save to repo
ggsave(
  filename = "Median_illnesses_by_season.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

Most outbreaks happen in winter, then spring, fall, and least in summer.

### b. Monthly Decomposition

**Plot of total illnesses over time (log transformed)**

```{r}
# 1. Sum of illnesses per year
yearly <- clean_df %>%
  group_by(Year) %>%
  summarise(total_illnesses = sum(Illnesses, na.rm = TRUE)) %>%
  ungroup()

# 2. Plot (log scale)
ggplot(yearly, aes(x = Year, y = log1p(total_illnesses))) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Total Illnesses by Year (log scale)",
    x = "Year",
    y = "log(Total Illnesses + 1)"
  )

# 3. Save to repo
ggsave(
  filename = "Log_total_illnesses_by_year.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

Clearly, there's an outlier; let's identify it:

```{r}
# 1. Filter all outbreaks from 1993-03
march_1993 <- clean_df %>%
  filter(Year == 1993, Month == 3)

# 2. See summary of illnesses and key variables
march_1993 %>%
  summarise(
    total_illnesses = sum(Illnesses, na.rm = TRUE),
    n_outbreaks = n(),
    states_involved = n_distinct(State_abbr),
    top_state = State_abbr[which.max(Illnesses)],
    primary_mode = names(sort(table(Primary_mode), decreasing = TRUE))[1]
  )

# 3. Identify which state it occurred in
march_1993 %>%
  group_by(State_abbr, Etiology) %>%
  summarise(
    total_ill = sum(Illnesses, na.rm = TRUE),
    hosp_per_ill = sum(Hosp_per_ill),
    deaths_per_ill = sum(Death_per_ill)) %>%
  arrange(desc(total_ill)) %>%
  slice_head(n = 10)
```

This date 1993-03-01 had 403,114 illness cases. According to a google search, this incident was caused by the 1993 Milwaukee Cryptosporidium waterborne outbreak, which sickened over 400,000 people — the largest waterborne disease outbreak in U.S. history.

**Plot of total hospitalizations over time (log transformed)**

```{r}
# 1. Sum of hospitalizations per year
yearly <- clean_df %>%
  group_by(Year) %>%
  summarise(total_hosp = sum(Hospitalizations, na.rm = TRUE)) %>%
  ungroup()

# 2. Plot (log scale)
ggplot(yearly, aes(x = Year, y = log1p(total_hosp))) +
  geom_line(color = "orange", linewidth = 1) +
  geom_point(color = "orange", size = 2) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Total Hospitalizations by Year (log scale)",
    x = "Year",
    y = "log(Total Hosp + 1)"
  )

# 3. Save to repo
ggsave(
  filename = "Log_total_hosp_by_year.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

**Plot of total deaths over time (log transformed)**

```{r}
# 1. Sum of hospitalizations per year
yearly <- clean_df %>%
  group_by(Year) %>%
  summarise(total_deaths = sum(Deaths, na.rm = TRUE)) %>%
  ungroup()

# 2. Plot (log scale)
ggplot(yearly, aes(x = Year, y = log1p(total_deaths))) +
  geom_line(color = "red", linewidth = 1) +
  geom_point(color = "red", size = 2) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Total Deaths by Year (log scale)",
    x = "Year",
    y = "log(Total Deaths + 1)"
  )

# 3. Save to repo
ggsave(
  filename = "Log_total_deaths_by_year.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

While total hospitalizations and deaths have increased over time, this may be due to increasing population size. So, it may be better to assess the ratio of hospitalizations/illnesses and deaths/illnesses over time.

**Plot of hospitalizations per illnesses over time**

```{r}
# 1. Monthly tends in hospitalizations per illnesses
monthly_trends <- clean_df %>%
  group_by(Year, Month) %>%
  summarise(total_hosp_per_ill = sum(Hosp_per_ill, na.rm = TRUE)) %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-")))

# 2. Plot of total illnesses 
ggplot(monthly_trends, aes(x = Date, y = total_hosp_per_ill)) +
  geom_line() +
  theme_minimal() +
  geom_smooth() +
  labs(title = "Monthly Trends in Total Hospitalizations/Illnesses")

# 3. Save to repo
ggsave(
  filename = "Monthly_trend_total_hosp_per_ill.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

**Plot of deaths per illnesses over time**

```{r}
# 1. Monthly tends in deaths per illnesses
monthly_trends <- clean_df %>%
  group_by(Year, Month) %>%
  summarise(total_death_per_ill = sum(Death_per_ill, na.rm = TRUE)) %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-")))

# 2. Plot of total illnesses 
ggplot(monthly_trends, aes(x = Date, y = total_death_per_ill)) +
  geom_line() +
  theme_minimal() +
  geom_smooth() +
  labs(title = "Monthly Trends in Total Deaths/Illnesses")

# 3. Save to repo
ggsave(
  filename = "Monthly_trend_total_death_per_ill.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

It seems like illnesses are becoming more severe over time, especially since 2000. Or, this could be reflecting an increased rate or improved system for reporting these metrics.
