---
title: "DSE 511 Final Project: Exploratory Data Analysis"
author: "Cecilia Albert-Black"
date: "2025-11-07"
output: html_document
---

Set working directory to R code folder

```{r}
setwd("C:/Users/q4q/repos/DSE511_Final_Project/Step 2. Exploratory Data Analysis/R Code")
```

Load necessary libraries

```{r}
library(readr)
library(tidyverse)
library(skimr)
library(scales)
```

## Exploratory Data Analysis of Clean Dataset

Import dataset from Step 1. Data Cleaning

```{r}
clean_df <- read_csv("C:/Users/q4q/repos/DSE511_Final_Project/Step 1. Data Cleaning/R code/outbreak_clean_r_v2.csv")

head(clean_df)
```

## Basic Dataset Overview

Check structure, missing data, and basic info.

```{r}
str(clean_df)
```

Summary statistics

```{r}
skim(clean_df)
```

Check missing values again

```{r}
missing_summary <- data.frame(
  missing_values = sort(colSums(is.na(clean_df)), decreasing = TRUE)
)
missing_summary
```

## Numeric Variable Distributions

Explore the distributions of numeric variables

```{r}
# 1. Define numeric variables
num_vars <- c(
  "Year", "Month",
  "Illnesses", "Hospitalizations", "Deaths",
  "Illnesses_std", "Hospitalizations_std", "Deaths_std",
  "Hosp_per_ill", "Death_per_ill"
)

# 2. Reshape the numeric data for plotting
num_vars_long <- clean_df %>%
  select(all_of(num_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

# 3. Plot histograms for each numeric variable
ggplot(num_vars_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Histograms of Numeric Variables",
    x = "Value",
    y = "Frequency"
  )

# 4. Save to repo
ggsave(
  filename = "Histogram_numeric_vars.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

It appears that all except Month and Year might need to be log-transformed to achieve are more normal distribution

```{r}
# 1. Define variables to log-transform
vars_to_log <- c(
  "Illnesses", "Hospitalizations", "Deaths",
  "Illnesses_std", "Hospitalizations_std", "Deaths_std",
  "Hosp_per_ill", "Death_per_ill"
)

# 2. Apply log transformation (add small constant to handle zeros)
clean_df <- clean_df %>%
  mutate(
    across(
      all_of(vars_to_log),
      ~ log(. + 1),                # add 1 to avoid log(0)
      .names = "log_{.col}"        # create new columns prefixed with 'log_'
    )
  )

# 3. Reshape data for easy plotting
log_vars_long <- clean_df %>%
  select(starts_with("log_")) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

# 4. Plot histograms for each log variable and add a kernel density estimation
ggplot(log_vars_long, aes(x = value)) +
  # histogram as density
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "steelblue", color = "white", alpha = 0.6) +
  # kernel density estimate
  geom_density(linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Histograms and Kernel Density Estimates of Log-Transformed Variables",
    x = "Log-transformed value",
    y = "Density"
  )
# 5. Save to repo
ggsave(
  filename = "Histograms_and_KDE_of_log_vars.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

Interpretation:

-   Most outbreaks are small, with few hospitalizations or deaths.

-   Log scale reveals the shape of the bulk of the data without extreme outbreaks dominating the plot.

-   There is a long tail of very large outbreaks for illnesses, which may represent epidemics or reporting anomalies.

## Seasonal or Yearly Trends

Show total illnesses per year to identify patterns in log-transformed variables.

```{r}
# 1. Sum of illnesses per year
yearly <- clean_df %>%
  group_by(Year) %>%
  summarise(total_illnesses = sum(Illnesses, na.rm = TRUE)) %>%
  ungroup()

# 2. Plot (log scale)
ggplot(yearly, aes(x = Year, y = log1p(total_illnesses))) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Total Illnesses by Year (log scale)",
    x = "Year",
    y = "log(Total Illnesses + 1)"
  )

# 3. Save to repo
ggsave(
  filename = "Total_illnesses_by_year.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

Interpretation:

-   Log scale highlights yearly trends without being dominated by extreme outbreaks.

-   Can reveal reporting improvements over time and identify peak years of large outbreaks.

Examine how outbreak severity changes by season

```{r}
# 1. Median illnesses by season
seasonal <- clean_df %>%
  group_by(Season) %>%
  summarise(median_illnesses = median(Illnesses, na.rm = TRUE)) %>%
  ungroup()

# 2. Plot
ggplot(seasonal, aes(x = Season, y = median_illnesses, fill = Season)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Median Illnesses by Season",
    x = "Season",
    y = "Median Illnesses"
  )

# 3. Save to repo
ggsave(
  filename = "Median_illnesses_by_season.png",  # file name
  width = 8, height = 5, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

## Top Contributing Factors

Identify top 50 etiologies (causes) contributing to outbreaks

```{r}
# 1. Isolate top 50 etiologies by total illnesses
top_etiology <- clean_df %>%
  group_by(Etiology) %>%
  summarise(total_illnesses = sum(Illnesses, na.rm = TRUE)) %>%
  arrange(desc(total_illnesses)) %>%
  slice_head(n = 50)

# 2. Plot
ggplot(top_etiology,
       aes(y = reorder(Etiology, total_illnesses),
           x = total_illnesses,
           fill = Etiology)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "mako") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 10 Etiologies by Total Illnesses",
    x = "Total Illnesses",
    y = "Etiology"
  )

# 3. Save to repo
ggsave(
  filename = "Top_etiologies_by_total_illnesses.png",  # file name
  width = 12, height = 12, units = "in",          # size in inches
  dpi = 300                                     # resolution for publication-quality image
)
```

Identify top 50 settings contributing to outbreaks

```{r}
# Isolate top 50 settings by total illnesses
top_setting <- clean_df %>%
  group_by(Setting) %>%
  summarise(total_illnesses = sum(Illnesses, na.rm = TRUE)) %>%
  arrange(desc(total_illnesses)) %>%
  slice_head(n = 50)

# 2. Plot
ggplot(top_setting,
       aes(y = reorder(Setting, total_illnesses),
           x = total_illnesses,
           fill = Setting)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "C") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 50 Settings by Total Illnesses",
    x = "Total Illnesses",
    y = "Setting"
  )

# 3. Save to repo
ggsave(
  filename = "Top_settings_by_total_illnesses.png",
  width = 12, height = 12, units = "in", dpi = 200
)
```

Interpretation:

-   Highlights primary sources of outbreaks.

-   Etiologies like Salmonella or Norovirus may dominate total illnesses.

-   Settings like long-term care facilities can explain outbreak clustering.

-   Using total illnesses (or log scale if desired) shows which factors drive high-impact outbreaks.

## Severity Ratios

Compare median hospitalization ratios by etiology

```{r}
# 1. Calculate median severity ratios by etiology 
severity <- clean_df %>%
  group_by(Etiology) %>%
  summarise(
    median_hosp_per_ill = median(Hosp_per_ill, na.rm = TRUE)
  ) %>%
  arrange(desc(median_hosp_per_ill)) %>%
  slice_head(n = 50)  

# 2. Clean up long labels for plotting 
library(stringr)

severity <- severity %>%
  mutate(Etiology_lab = str_wrap(Etiology, width = 35))  # wrap long names

# 3. Make the plot object
ggplot(
  severity,
  aes(
    x = median_hosp_per_ill,
    y = reorder(Etiology_lab, median_hosp_per_ill),
    fill = Etiology
  )
) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "mako") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 10 Etiologies by Median Hospitalizations per Illness",
    x = "Median Hospitalizations per Illness",
    y = "Etiology"
  )


# 4. Save to file
ggsave(
  filename = "Top_etiologies_by_median_hospitalizations_per_illness.png", 
  width = 12, height = 12,
  units = "in",
  dpi = 300
)
```

Interpretation:

-   Median ratios are more robust to extreme cases than the mean.

-   Etiologies with higher hospitalization or death rates may indicate more severe outbreaks, even if total outbreak counts are lower.

## Deaths by Etiology

Total deaths by etiology

```{r}
# 1. Compute total deaths per etiology
deaths_total <- clean_df %>%
  group_by(Etiology) %>%
  summarise(
    total_deaths = sum(Deaths, na.rm = TRUE),
    total_illnesses = sum(Illnesses, na.rm = TRUE),
    n_outbreaks = n()
  ) %>%
  filter(total_deaths > 0) %>%        
  arrange(desc(total_deaths)) %>%
  slice_head(n = 50)                  

# 2. Plot
ggplot(deaths_total,
       aes(y = reorder(Etiology, total_deaths),
           x = total_deaths,
           fill = Etiology)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "plasma") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 50 Etiology  by Total Deaths",
    x = "Total deaths (sum across all outbreaks)",
    y = "Etiology"
  )


# 3. Save to repo
ggsave(
  filename = "Top_etiology_by_total_deaths.png",
  width = 12, height = 12, units = "in", dpi = 300
)
```

Mean deaths per outbreak: highlights average lethality per outbreak

```{r}
# 1. Compute mean deaths per etiology
deaths_mean <- clean_df %>%
  group_by(Etiology) %>%
  summarise(
    n_outbreaks = n(),
    mean_deaths = mean(Deaths, na.rm = TRUE)
  ) %>%
  filter(n_outbreaks >= 50) %>%
  arrange(desc(mean_deaths)) %>%
  slice_head(n = 50)

# 2. Plot
ggplot(deaths_mean,
       aes(y = reorder(Etiology, mean_deaths),
           x = mean_deaths,
           fill = Etiology)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "rocket") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 50 Etiology Groups by Average Deaths per Outbreak",
    x = "Average deaths per outbreak",
    y = "Etiology group"
  )

# 3. Save to repo
ggsave(
  filename = "Top_etiology_by_mean_deaths.png",
  width = 12, height = 12, units = "in", dpi = 300
)
```

This metric balances outbreak count and lethality. A group with few but deadly outbreaks (e.g., *Listeria*) can stand out compared to frequent but mild ones (e.g., *Norovirus*).

## Correlation Heatmap

```{r}
# 1. Call out variables
corr_cols <- c("Illnesses", "Hospitalizations", "Deaths", "Hosp_per_ill", "Death_per_ill")

# 2. Compute correlation matrix
corr_matrix <- clean_df %>%
  select(all_of(corr_cols)) %>%
  cor(use = "pairwise.complete.obs") %>%
  as.data.frame()

# 3. Convert to long format for ggplot
corr_long <- corr_matrix %>%
  mutate(var1 = rownames(corr_matrix)) %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")

# 4. Plot
ggplot(corr_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), color = "black", size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "darkred", midpoint = 0) +
  coord_equal() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Correlation Heatmap of Key Variables",
    x = NULL,
    y = NULL
  )

# 5. Save to repo
ggsave(
  filename = "Correlation_heatmap_num_vars.png",
  width = 8, height = 8, units = "in", dpi = 300
)
```

Strong correlations can indicate redundant variables; none are identified are highly correlated.

## Spatial Trends

Median illness counts by state (across all years)

```{r}
# 1. Compute median illnesses per state
geo <- clean_df %>%
  group_by(State_abbr) %>%
  summarise(median_illnesses = median(Illnesses, na.rm = TRUE)) %>%
  arrange(desc(median_illnesses)) %>%
  slice_head(n = 15)

# 2. Plot top 15 states by median illnesses
ggplot(
  geo,
  aes(
    x = reorder(State_abbr, -median_illnesses),
    y = median_illnesses,
    fill = State_abbr
  )
) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 15 States by Median Number of Illnesses",
    x = "State",
    y = "Median Illnesses"
  )

# 3. Save the plot
ggsave(
  filename = "Top_15_States_by_Median_Illnesses.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

Median yearly illnesses by state

```{r}
# 1. Compute total illnesses per state-year, then median yearly illness rate per state
yearly_geo <- clean_df %>%
  group_by(State_abbr, Year) %>%
  summarise(yearly_illnesses = sum(Illnesses, na.rm = TRUE), .groups = "drop") %>%
  group_by(State_abbr) %>%
  summarise(median_yearly_illnesses = median(yearly_illnesses, na.rm = TRUE)) %>%
  arrange(desc(median_yearly_illnesses)) %>%
  slice_head(n = 15)

# 2. Plot top 15 states by median yearly illness rate
ggplot(
  yearly_geo,
  aes(
    x = reorder(State_abbr, -median_yearly_illnesses),
    y = median_yearly_illnesses,
    fill = State_abbr
  )
) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 15 States by Median Yearly Illness Rate",
    x = "State",
    y = "Median Yearly Illnesses"
  )

# 3. Save the plot
ggsave(
  filename = "Top_15_States_by_Median_Yearly_Illnesses.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

Median yearly hospitalizations by state

```{r}
# 1. Compute total hospitalizations per state-year, then median yearly hospitalizations per state
yearly_hosp <- clean_df %>%
  group_by(State_abbr, Year) %>%
  summarise(yearly_hospitalizations = sum(Hospitalizations, na.rm = TRUE), .groups = "drop") %>%
  group_by(State_abbr) %>%
  summarise(median_yearly_hospitalizations = median(yearly_hospitalizations, na.rm = TRUE)) %>%
  arrange(desc(median_yearly_hospitalizations)) %>%
  slice_head(n = 15)

# 2. Plot top 15 states by median yearly hospitalizations
ggplot(
  yearly_hosp,
  aes(
    x = reorder(State_abbr, -median_yearly_hospitalizations),
    y = median_yearly_hospitalizations,
    fill = State_abbr
  )
) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 15 States by Median Yearly Hospitalizations",
    x = "State",
    y = "Median Yearly Hospitalizations"
  )

# 3. Save the plot
ggsave(
  filename = "Top_15_States_by_Median_Yearly_Hospitalizations.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

Median yearly deaths rate by state

```{r}
# 1. Compute total deaths per state-year, then median yearly deaths per state
yearly_deaths <- clean_df %>%
  group_by(State_abbr, Year) %>%
  summarise(yearly_deaths = sum(Deaths, na.rm = TRUE), .groups = "drop") %>%
  group_by(State_abbr) %>%
  summarise(median_yearly_deaths = median(yearly_deaths, na.rm = TRUE)) %>%
  arrange(desc(median_yearly_deaths)) %>%
  slice_head(n = 15)

# 2. Plot top 15 states by median yearly deaths
ggplot(
  yearly_deaths,
  aes(
    x = reorder(State_abbr, -median_yearly_deaths),
    y = median_yearly_deaths,
    fill = State_abbr
  )
) +
  geom_col(show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top 15 States by Median Yearly Deaths",
    x = "State",
    y = "Median Yearly Deaths"
  )

# 3. Save the plot
ggsave(
  filename = "Top_15_States_by_Median_Yearly_Deaths.png",
  width = 10, height = 5, units = "in", dpi = 300
)
```

It appears that outbreak severity is most pronounced in areas outside of the major 51 US states. Maybe we'll consider omitting these areas, since we have limited information on them. However, this will introduce a bias in our dataset since we will not be considering US territories, native american reservations, and army bases.

## Temporal Trends
