---
title: "DSE 511 Final Project Machine Learning in R"
author: "Cecilia Albert-Black"
date: "2025-11-11"
output: html_document
---

Set working directory to R code folder

```{r}
setwd("C:/Users/q4q/repos/DSE511_Final_Project/Step 4. Supervised Machine Learning/R Code")
```

Load necessary libraries

```{r}
library(readr)
library(tidyverse)
```

Import dataset from Step 1. Data Cleaning

```{r}
clean_df <- read_csv("C:/Users/q4q/repos/DSE511_Final_Project/Step 1. Data Cleaning/R code/outbreak_clean_r_v2.csv")

head(clean_df)
```

## Defining the Objective

We want to model Illnesses, Hospitalizations, and Deaths, which are:

-   Count-based (non-negative, right-skewed)

-   Highly overdispersed (variance \>\> mean)

-   Driven by categorical predictors (Etiology, Setting, Season, etc.)

-   Complex, nonlinear, and potentially interactive (e.g., Setting × Mode interactions)

Because of these confounding complexities, no single ML type that is “perfect.”

We will be comparing 3 different models:

1.  Multiple Linear Regression (baseline model)
2.  Random Forest (robust nonlinear ML model)
3.  Gradient Boosting (Precision-focused ML model)

## Modeling
